---
title: "Egyptian Skull Sizes over Time"
description: |
  An attempt at an application of the Manova. The main contention with this problem is the overwhelming lack of linearity between pairs of dependent variables. 

base_url: https://github.com/gavandrewj/myBlog/

categories:
  - Manova
  - Independent Samples   
  
author:
  - name: Gavin Gordon
    url: https://example.com/norajones
date: 01-05-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,cache = T,message = F,warning = F)
```

```{r}
library(tidyverse)
library(fitdistrplus)
library(SHT)
library(vroom)
library(plotly)
library(gt)
library(gtsummary)
library(Hotelling)
library(pander)
library(expss)
library(beanplot)
library(pairwiseCI)
library(heplots)
library(GGally)
library(Hmisc)
library(ellipse)
library(RColorBrewer)
```
# Problem (Question 6.18)
This problem is taken from **Johnson 1992**. Researchers have suggested that the change in skull size over time is evidence of the interbreeding of a resident population with immigrant populations. Four measurements were made of male Egyptian skulls over five time periods. The aim of this test is to determine if there is a difference in the skull sizes over time. 

```{r}
#Importing data
data("Skulls")
data <- Skulls
rm(Skulls)

data$epoch <- factor(data$epoch)
#var.labels <- c("Epoch","MaxBreath","BasHeight","Baslength","NasHeight","Outliers")
#data$loglength <- log(data$length)
#data$logwidth <- log(data$width)
#data$logheight <- log(data$height)
```

```{r }
data$madist <- Mahalanobis(data[,2:5])
#summary( Mahalanobis(data[,2:5]))
#plot(density(Mahalanobis(data[,2:5])))
data$pout <- factor(ifelse(data$madist > 9 ,1,0))


```

We will start off by taking a look at outliers. Honestly I don't know what to make of this dataset when it comes to applying the Manova. There is supposed to be linearity effects between the dependent variable pairs.

<center>

```{r fig.align='center'}
c4000BC <- data %>% filter(epoch == "c4000BC")
c4000BC$madist <- Mahalanobis(c4000BC[,2:5])
#summary( Mahalanobis(data[,2:5]))
#plot(density(Mahalanobis(c4000BC[,2:5])))
c4000BC$pout <- factor(ifelse(c4000BC$madist > 7 ,1,0))
scatterc4000 <- c4000BC %>% select(!c(epoch,madist)) %>% ggpairs(aes(color = pout),columnLabels = c("Maxbreath",
                                                                                                    "BasHeight",
                                                                                                    "BasLength",
                                                                                                    "NasHeight",
                                                                                                    "Outliers")) + theme_bw()

ggplotly(scatterc4000, height = 600, width=1000)

```

</center>





<center>

```{r fig.align='center'}
c3300BC <- data %>% filter(epoch == "c3300BC")
c3300BC$madist <- Mahalanobis(c3300BC[,2:5])
#summary( Mahalanobis(data[,2:5]))
#plot(density(Mahalanobis(c3300BC[,2:5])))
c3300BC$pout <- factor(ifelse(c3300BC$madist > 7 ,1,0))
scatterc3300 <- c3300BC %>% select(!c(epoch,madist)) %>% ggpairs(aes(color = pout),columnLabels = c("Maxbreath",
                                                                                                    "BasHeight",
                                                                                                    "BasLength",
                                                                                                    "NasHeight",
                                                                                                    "Outliers")) + theme_bw()

ggplotly(scatterc3300, height = 600, width=1000)

```

</center>



<center>

```{r fig.align='center'}
c1850BC <- data %>% filter(epoch == "c1850BC")
c1850BC$madist <- Mahalanobis(c1850BC[,2:5])
#summary( Mahalanobis(data[,2:5]))
#plot(density(Mahalanobis(c1850BC[,2:5])))
c1850BC$pout <- factor(ifelse(c1850BC$madist > 7 ,1,0))
scatterc1850 <- c1850BC %>% select(!c(epoch,madist)) %>% ggpairs(aes(color = pout),columnLabels = c("Maxbreath",
                                                                                                    "BasHeight",
                                                                                                    "BasLength",
                                                                                                    "NasHeight",
                                                                                                    "Outliers")) + theme_bw()

ggplotly(scatterc1850, height = 600, width=1000)

```

</center>



<center>

```{r fig.align='center'}
c200BC <- data %>% filter(epoch == "c200BC")
c200BC$madist <- Mahalanobis(c200BC[,2:5])
#summary( Mahalanobis(data[,2:5]))
#plot(density(Mahalanobis(c200BC[,2:5])))
c200BC$pout <- factor(ifelse(c200BC$madist > 7 ,1,0))
scatterc200 <- c200BC %>% select(!c(epoch,madist)) %>% ggpairs(aes(color = pout),columnLabels = c("Maxbreath",
                                                                                                    "BasHeight",
                                                                                                    "BasLength",
                                                                                                    "NasHeight",
                                                                                                    "Outliers")) + theme_bw()

ggplotly(scatterc200, height = 600, width=1000)

```

</center>


<center>

```{r fig.align='center'}
cAD150 <- data %>% filter(epoch == "cAD150")
cAD150$madist <- Mahalanobis(cAD150[,2:5])
#summary( Mahalanobis(data[,2:5]))
#plot(density(Mahalanobis(cAD150[,2:5])))
cAD150$pout <- factor(ifelse(cAD150$madist > 7 ,1,0))
scatterc150 <- cAD150 %>% select(!c(epoch,madist)) %>% ggpairs(aes(color = pout),columnLabels = c("Maxbreath",
                                                                                                    "BasHeight",
                                                                                                    "BasLength",
                                                                                                    "NasHeight",
                                                                                                    "Outliers")) + theme_bw()

ggplotly(scatterc150, height = 600, width=1000)

```

</center>

## References

Johnson, R. A., & Wichern, D. W. (1992). Applied multivariate statistical analysis. Englewood Cliffs, N.J: Prentice Hall  


NCSS Hotelling’S Twosample T2. [pdf] Available at: <https://ncss-wpengine.netdna-ssl.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Hotellings_Two-Sample_T2.pdf> 
