---
title: "Perceived Working Environments"
description: |
  A short description of the post.

base_url: https://github.com/gavandrewj/myBlog/

categories:
  - Factor Analysis
  - Multivariate Analysis
  
  
author:
  - name: Gavin Gordon
    url: https://example.com/norajones
date: 01-19-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = F, message = F,cache = T)
```

```{r}
library(nFactors)
library(GPArotation)
library(ggcorrplot)
library(corrplot)
library(haven)
library(tidyverse)
library(semPlot)
library(gplots)
library(RColorBrewer)
library(caret)
```

#factor analysis presentation
#Factors are regarded as latent variables that cannot be observed directly, but are imperfectly
#assessed through their relationship to other variables.
#find the degree to which latent, composite factors account for the observed variance
#of those manifest variables
#interpretable in terms of the manifest variables
#data reduction
# noise culling
# question item reduction

#Load the R libraries

```{r}
data <- data.frame(read_spss("C:/Users/gavin/Desktop/Projects/factor_analysis/data.sav"))

toanalyze <- data[,1:15]

# Compare the case to variables
nrow(toanalyze)/ncol(toanalyze)
```

```{r}
#Initial look at the correlations

datacor <- cor(toanalyze)
corrplot(datacor, order = "hclust")

```


```{r}
highCorr <- findCorrelation(datacor, cutoff = .3)
length(highCorr)
filtereddata <- toanalyze[, highCorr]
```

# check the suggested number of factors

```{r}
plot(nScree(toanalyze))

nScree(toanalyze)

plotuScree(toanalyze)
```

```{r}
#calculate the eigen values
eigenvalues <- eigen(cor(toanalyze))$values 

# how many are greater than one
length(eigenvalues[eigenvalues > 1])

#Let's take a look at them
eigenvalues[eigenvalues > 1]
```

```{r}
factanal(toanalyze , factors =4)

```

```{r}
factanal(toanalyze , factors =9)

```
# rotations

```{r}
output <-  factanal(toanalyze, factors =4, rotation="oblimin")
output

```

```{r}

heatmap.2(output$loadings,
           col=brewer.pal(9, "Greens"),
           trace="none",
           key=FALSE,
           dend="none",
           Colv=FALSE,
           cexCol = 1.2,
           main="\n\n\n\n\nFactor loadings for brand adjectives")

```

```{r}
semPaths(output , what="est" , residuals=FALSE ,
         cut =0.3 , posCol=c("white" , "darkgreen") , negCol=c("white" , "red"),
         edge.label.cex =0.75 , nCharNodes =7)

# Second Iteration
#potential for being dropped
# 3,7,14,15,23
```

```{r}
toanalyze <- toanalyze %>% dplyr::select(!how.satisfy3)
nScree(toanalyze)

#calculate the eigen values
eigenvalues <- eigen(cor(toanalyze))$values 

# how many are greater than one
length(eigenvalues[eigenvalues > 1])

#Let's take a look at them
eigenvalues[eigenvalues > 1]



factanal(toanalyze , factors =7)


factanal(toanalyze , factors =8)


output <-  factanal(toanalyze, factors =7, rotation="oblimin")

heatmap.2(output$loadings,
          col=brewer.pal(9, "Greens"),
          trace="none",
          key=FALSE,
          dend="none",
          Colv=FALSE,
          cexCol = 1.2,
          main="\n\n\n\n\nFactor loadings for brand adjectives")


semPaths(output , what="est" , residuals=FALSE ,
         cut =0.3 , posCol=c("white" , "darkgreen") , negCol=c("white" , "red"),
         edge.label.cex =0.75 , nCharNodes =7)
```

```{r}
# third iteration
#potential for being dropped
#7,15,23

toanalyze <- toanalyze %>% dplyr::select(!how.satisfy7)
nScree(toanalyze)

#calculate the eigen values
eigenvalues <- eigen(cor(toanalyze))$values 

# how many are greater than one
length(eigenvalues[eigenvalues > 1])

#Let's take a look at them
eigenvalues[eigenvalues > 1]



factanal(toanalyze , factors =7)




output <-  factanal(toanalyze, factors =7, rotation="oblimin")
output

heatmap.2(output$loadings,
          col=brewer.pal(9, "Greens"),
          trace="none",
          key=FALSE,
          dend="none",
          Colv=FALSE,
          cexCol = 1.2,
          main="\n\n\n\n\nFactor loadings for brand adjectives")


semPaths(output , what="est" , residuals=FALSE ,
         cut =0.3 , posCol=c("white" , "darkgreen") , negCol=c("white" , "red"),
         edge.label.cex =0.75 , nCharNodes =7)

```

# fourth iteration
potential for being dropped  
14,15,23

```{r}
toanalyze <- toanalyze %>% dplyr::select(!how.satisfy15)
nScree(toanalyze)

#calculate the eigen values
eigenvalues <- eigen(cor(toanalyze))$values 

# how many are greater than one
length(eigenvalues[eigenvalues > 1])

#Let's take a look at them
eigenvalues[eigenvalues > 1]



factanal(toanalyze , factors =7)




output <-  factanal(toanalyze, factors =7, rotation="oblimin")
output

heatmap.2(output$loadings,
          col=brewer.pal(9, "Greens"),
          trace="none",
          key=FALSE,
          dend="none",
          Colv=FALSE,
          cexCol = 1.2,
          main="\n\n\n\n\nFactor loadings for brand adjectives")


semPaths(output , what="est" , residuals=FALSE ,
         cut =0.3 , posCol=c("white" , "darkgreen") , negCol=c("white" , "red"),
         edge.label.cex =0.75 , nCharNodes =7)
```

#fifth iteration
potential for being dropped  
23
```{r}
toanalyze <- toanalyze %>% dplyr::select(!how.satisfy14)
nScree(toanalyze)

#calculate the eigen values
eigenvalues <- eigen(cor(toanalyze))$values 

# how many are greater than one
length(eigenvalues[eigenvalues > 1])

#Let's take a look at them
eigenvalues[eigenvalues > 1]



factanal(toanalyze , factors =7)




output <-  factanal(toanalyze, factors =7, rotation="oblimin")
output

heatmap.2(output$loadings,
          col=brewer.pal(9, "Greens"),
          trace="none",
          key=FALSE,
          dend="none",
          Colv=FALSE,
          cexCol = 1.2,
          main="\n\n\n\n\nFactor loadings for brand adjectives")


semPaths(output , what="est" , residuals=FALSE ,
         cut =0.3 , posCol=c("white" , "darkgreen") , negCol=c("white" , "red"),
         edge.label.cex =0.75 , nCharNodes =7)

```

#fifth iteration
potential for being dropped
23

```{r}
toanalyze <- toanalyze %>% dplyr::select(!how.satisfy7)
toanalyze <- toanalyze %>% dplyr::select(!how.satisfy15)

nScree(toanalyze)

#calculate the eigen values
eigenvalues <- eigen(cor(toanalyze))$values 

# how many are greater than one
length(eigenvalues[eigenvalues > 1])

#Let's take a look at them
eigenvalues[eigenvalues > 1]



factanal(toanalyze , factors =7)




output <-  factanal(toanalyze, factors =4, rotation="oblimin")
output

heatmap.2(output$loadings,
          col=brewer.pal(9, "Greens"),
          trace="none",
          key=FALSE,
          dend="none",
          Colv=FALSE,
          cexCol = 1.2,
          main="\n\n\n\n\nFactor loadings for brand adjectives")


semPaths(output , what="est" , residuals=FALSE ,
         cut =0.3 , posCol=c("white" , "darkgreen") , negCol=c("white" , "red"),
         edge.label.cex =0.75 , nCharNodes =7)

```





## Only Variable with high corr (>0.3)

```{r}
nScree(filtereddata)

#calculate the eigen values
eigenvalues <- eigen(cor(filtereddata))$values 

# how many are greater than one
length(eigenvalues[eigenvalues > 1])

#Let's take a look at them
eigenvalues[eigenvalues > 1]



factanal(filtereddata , factors =3)




output <-  factanal(filtereddata, factors =3, rotation="oblimin")
output

heatmap.2(output$loadings,
          col=brewer.pal(9, "Greens"),
          trace="none",
          key=FALSE,
          dend="none",
          Colv=FALSE,
          cexCol = 1.2,
          main="\n\n\n\n\nFactor loadings for brand adjectives")


semPaths(output , what="est" , residuals=FALSE ,
         cut =0.3 , posCol=c("white" , "darkgreen") , negCol=c("white" , "red"),
         edge.label.cex =0.75 , nCharNodes =7)

```
