---
title: "Egyptian Skull Sizes over Time"
description: |
  An attempt at an application of the Manova. The main contention with this problem is the overwhelming lack of linearity between pairs of dependent variables. 

base_url: https://github.com/gavandrewj/myBlog/

categories:
  - Manova
  - Independent Samples   
  
author:
  - name: Gavin Gordon
    url: https://example.com/norajones
date: 01-05-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,cache = T,message = F,warning = F)
```

```{r}
library(tidyverse)
library(fitdistrplus)
library(SHT)
library(vroom)
library(plotly)
library(gt)
library(gtsummary)
library(Hotelling)
library(pander)
library(expss)
library(beanplot)
library(pairwiseCI)
library(heplots)
library(GGally)
library(Hmisc)
library(ellipse)
library(RColorBrewer)
```
# Problem (Question 6.18)
This problem is taken from **Johnson 1992**. Researchers have suggested that the change in skull size over time is evidence of the interbreeding of a resident population with immigrant populations. Four measurements were made of male Egyptian skulls over five time periods. The aim of this test is to determine if there is a difference in the skull sizes over time. 

```{r}
#Importing data
data("Skulls")
data <- Skulls
rm(Skulls)

data$epoch <- factor(data$epoch)
#var.labels <- c("Epoch","MaxBreath","BasHeight","Baslength","NasHeight","Outliers")
#data$loglength <- log(data$length)
#data$logwidth <- log(data$width)
#data$logheight <- log(data$height)
```

```{r }
data$madist <- Mahalanobis(data[,2:5])
#summary( Mahalanobis(data[,2:5]))
#plot(density(Mahalanobis(data[,2:5])))
data$pout <- factor(data$madist > 9)

```

We will start off by taking a look at outliers. Honestly I don't know what to make of this dataset when it comes to applying the Manova. 

```{r}
my_colors <- brewer.pal(5, "Spectral")
my_colors <- colorRampPalette(my_colors)(100)

seecor <- data  %>% select(!c(madist,pout)) %>%
  filter(epoch == "c4000BC") %>% select(!epoch) %>% cor()

ord <- order(seecor[1, ])
data_ord <- seecor[ord, ord]
plotcorr(data_ord , col=my_colors[data_ord*50+50] , mar=c(1,1,1,1)  )


```

```{r fig.align='center' , fig.cap= "Bivariate Plots for c4000BC Epoch"}
pl_colorscale = list(c(0.0, '#119dff'),
                  c(0.5, '#119dff'),
                  c(0.5, '#ef553b'),
                  c(1, '#ef553b'))

axis = list(showline=FALSE,
            zeroline=FALSE,
            gridcolor='#ffff',
            ticklen=4)

fig <- data %>% filter(epoch == "c4000BC") %>%
  plot_ly() 
fig <- fig %>%
  add_trace(
    type = 'splom',
    dimensions = list(
      list(label='MaxBreath', values=~mb),
      list(label='BasHeight', values=~bh),
      list(label='BasLength', values=~bl),
      list(label='NasHeight', values=~nh)
    ),
    text=~factor(pout, labels=c("non-diabetic","diabetic")),
    diagonal=list(visible=F),
    marker = list(
      color = ~pout,
      colorscale = pl_colorscale,
      size = 5,
      line = list(
        width = 1,
        color = 'rgb(230,230,230)'
      )
    )
  ) 

fig <- fig %>%
  layout(
    title= 'ScatterPlots in Epoch c4000BC',
    hovermode='closest',
    dragmode= 'select',
    plot_bgcolor='rgba(240,240,240, 0.95)',
    xaxis=list(domain=NULL, showline=F, zeroline=F, gridcolor='#ffff', ticklen=4),
    yaxis=list(domain=NULL, showline=F, zeroline=F, gridcolor='#ffff', ticklen=4),
    xaxis2=axis,
    xaxis3=axis,
    xaxis4=axis,
    yaxis2=axis,
    yaxis3=axis,
    yaxis4=axis
  )

fig %>% style(showupperhalf = F)

```




## References

Johnson, R. A., & Wichern, D. W. (1992). Applied multivariate statistical analysis. Englewood Cliffs, N.J: Prentice Hall  


NCSS Hotelling’S Twosample T2. [pdf] Available at: <https://ncss-wpengine.netdna-ssl.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Hotellings_Two-Sample_T2.pdf> 
