---
title: "Wine Analysis"
description: |
  This analysis explores an application of the Manova to measured chemical features of wine samples that use different cultivars in the brewing process. The analysis includes the use of a k sample Brehen Fisher test for high dimensional data; this does not require the equality of covariance matrices assumption to be satisfied. 
  
base_url: https://github.com/gavandrewj/myBlog/

categories:
  - Multivariate Analysis
  - Manova

author:
  - name: Gavin Gordon
    url: https://example.com/norajones
date: 01-23-2021
output:
  distill::distill_article:
    self_contained: false
---

# Introduction

It should be noted that there is little theory behind why the features were selected as the dependent measures. I am for the most part interested in an application of the Manova, thus through plots I found features that are separated for each of the tested categorical variable levels and ignored much of the other variables. For the same reason the number of variables was kept to three for illustrative purposes as well as to propose a question about the linearity assumption. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,cache = T,message = F,wwarning = F)
```

```{r}
library(tidyverse)
library(SHT)
library(vroom)
library(plotly)
library(gt)
library(gtsummary)
library(Hotelling)
library(pander)
library(expss)
library(beanplot)
library(pairwiseCI)
library(heplots)
library(GGally)
library(lubridate)
```

```{r}
names <- c("cultivar","alcohol","malic_acid","ash","alcalinity_of_ash" ,"magnesium" , "total_phenols" ,"flavanoids",
"nonflavanoid_phenols","proanthocyanins","color_intensity","hue","OD280/OD315_of_diluted_wines",
"proline") 
#data <- vroom("C:/Users/gavin/Documents/GitHub/blogdata/wine_data.csv",col_names = names)

dataset <- vroom("C:/Users/gavin/Documents/GitHub/blogdata/wine_data.csv",col_names = names)
rm(names)
dataset$cultivar <- factor(dataset$cultivar)
#data$cultivar <- factor(data$cultivar)

```

To start things off we will go straight to the 3D plot as this will give us the best view of what is happening with the data.
&nbsp;

```{r}
dataset <- dataset %>% select(cultivar,total_phenols,flavanoids,alcohol)

plot_ly(data = dataset, x = dataset$total_phenols,y = dataset$flavanoids,z = dataset$alcohol,color = dataset$cultivar) %>% layout(title = "Plot of Dependent Variables by Cultivar",
                scene = list(
                  xaxis = list(title = "Total Phenols"),
                  yaxis = list(title = "Flavanoids"),
                  zaxis = list(title = "Alcohol")
                  )
                )

outlier <- data.frame(total_phenols = 3.18,flavanoids = 5.08,alcohol = 11.56)
dataset <- anti_join(dataset,outlier)
```

My view of looking at this problem is to imagine (haven't figured out how to easily do this in plotly) an ellipsoid centered at the respective group means, that will aim to encompass the data points for each group. To the extent that each ellipsoid contains it's points and separates itself from other ellipsoids will determine whether or not there is a difference across the levels of the categorical variable (cultivar).  

We can see that there is likely much more variability in group 2 than other groups and there is overlap somewhat with groups 1 and 3. The opposite occurs for groups 1 and 2 as these are none overlapping groups. It is expected that the Manova will determine that there are group differences.

Also there is one point that stands out as an outlier which belongs to group 2. The level of influence this point has is not checked; it was just removed as it is at such a distance from any of the groups that it is more likely to be an error.


## Checking the Assumptions for the Manova

1. **Independence of Observations**  

If some of the samples were brewed in the same batch then this would introduce dependencies between the measurements; this is assumed to not have happened. 

2. **Normality and Outlier Analysis**  

The sample sizes within the Group levels would be enough to be considered a large sample (greater than or equal to 20) thus the Central Limit theorem would take effect to assure normality of the sample mean. The following 3D plots highlight points for examination (the mahalonobis distances are calculated and used to marked as outliers). I am not in a position to remove points other than those that clearly does not fit into the distributions.   

&nbsp;
```{r}

group1 <- dataset %>% filter(cultivar == 1) %>% select(alcohol,flavanoids,total_phenols) 
group1$dist <-  mahalanobis(group1,colMeans(group1),cov(group1)) >7


plot_ly(data = group1, x = group1$total_phenols,y = group1$flavanoids,z = group1$alcohol,color = group1$dist) %>% layout(title = "Plot of Potential Outliers for Cultivar One",
                scene = list(
                  xaxis = list(title = "Total Phenols"),
                  yaxis = list(title = "Flavanoids"),
                  zaxis = list(title = "Alcohol")
                  )
                )

group1$cultivar <- 1 
```
&nbsp;
```{r}

group2 <- dataset %>% filter(cultivar == 2) %>% select(alcohol,flavanoids,total_phenols) 
group2$dist <-  mahalanobis(group2,colMeans(group2),cov(group2)) >7


plot_ly(data = group2, x = group2$total_phenols,y = group2$flavanoids,z = group2$alcohol,color = group2$dist) %>% layout(title = "Plot of Potential Outliers for Cultivar Two",
                scene = list(
                  xaxis = list(title = "Total Phenols"),
                  yaxis = list(title = "Flavanoids"),
                  zaxis = list(title = "Alcohol")
                  )
                )
group2$cultivar <- 2

```
&nbsp;
```{r}

group3 <- dataset %>% filter(cultivar == 3) %>% select(alcohol,flavanoids,total_phenols) 
group3$dist <-  mahalanobis(group3,colMeans(group3),cov(group3)) >7


plot_ly(data = group3, x = group3$total_phenols,y = group3$flavanoids,z = group3$alcohol,color = group3$dist) %>% layout(title = "Plot of Potential Outliers for Cultivar Three",
                scene = list(
                  xaxis = list(title = "Total Phenols"),
                  yaxis = list(title = "Flavanoids"),
                  zaxis = list(title = "Alcohol")
                  )
                )

outlier <- data.frame(total_phenols = 2.8,flavanoids = 1.31,alcohol = 14.34)
group3 <- anti_join(group3,outlier)
group3$cultivar <- 3

dataset <- rbind(group1,group2,group3)
dataset <- dataset[,-4]
dataset$cultivar <- factor(dataset$cultivar)
```

The outlier for cultivar three that is labelled with coordinates (2.8,1.31,14.34) is removed.


3. **Linearity and Multicollinearity of Dependent Variables**  

So we expect that the dependent variables have a linear relationship and a lack thereof would decrease the power of the analysis. How this seems to work is that as linearity increases, the length of the minor axes of the ellipsoid begins to decrease while the length of the major axis increases so the points take up a more narrow region of space that stretches for longer. From the 3D scatter plot we can notice good separation of the groups; we see this behavior in cultivar one but less so in two, and even less so in three. Moreover we can see from the 2D scatterplots presented below, four of the nine within group correlations are practically zero. So  After what extent of non-linearity can we consider the power of this test to be so diminished that we default back to using separate Anovas? And of course how do we manage this when moving beyond three dependent variables?

```{r fig.align='center'}
#dataset <- dataset %>% select(cultivar,total_phenols,flavanoids,nonflavanoid_phenols)
#alcohol,color_intensity,total_phenols,flavanoids

#mg, fla, alkaash - good discrim not good corr

#phenol, non flav, non phen # pretty goof

#proline, alkaash

#color int, hue

bivar <- dataset %>% ggpairs(aes(color = cultivar),columnLabels = c("Total Phenols","Flavanoids","Alcohol","Cultivar")) + theme_bw() 

ggplotly(bivar,height = 700,width = 800)

```



4. **Equality of Variance**

This is not a requirement for the test that will be executed. 


## Analysis

As expected the p value from the test is below 0.05; therefore if we were to assume that the null hypothesis is true (that there is no difference between the group means) then the likelihood of the data that was observed, is very low. The data would be more consistent with the hypothesis that the means are different among the groups. 

&nbsp;
<center>
```{r}


## CRAN-purpose small example
tinylist = list()


mat <- subset(dataset,cultivar == 1)
mat <- matrix(c(mat$total_phenols,mat$flavanoids,mat$alcohol),ncol = 3)
tinylist[[1]] <- mat 

mat <- subset(dataset,cultivar == 2)
mat <- matrix(c(mat$total_phenols,mat$flavanoids,mat$alcohol),ncol = 3)
tinylist[[2]] <- mat 
 
mat <- subset(dataset,cultivar == 3)
mat <- matrix(c(mat$total_phenols,mat$flavanoids,mat$alcohol),ncol = 3)
tinylist[[3]] <- mat 

 
pander(meank.2009ZX(tinylist),caption =  "Test of difference of Cultivar group means")

```
</center>

We will now explore bonferonni confidence intervals to determine along which variables there are differences. 

<center>
```{r,out.width= '80%'}
means <- dataset %>% group_by(cultivar) %>% summarise(alcohol = mean(alcohol),total_phenols = mean(total_phenols),flavanoids = mean(flavanoids))
means$cultivar <-  factor(means$cultivar)

meanstp <- means %>% select(cultivar,total_phenols) %>% ggplot(aes(cultivar,total_phenols)) + 
  geom_point() + 
  geom_segment( aes(x=cultivar, xend=cultivar, y=0, yend=total_phenols)) + 
  theme_bw() + 
  labs(x = "Cultivar",y = "Mean Total Phenol Level",title = "Difference in Total Phenol Levels across Cultivars") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplotly(meanstp)
```
</center>
<center>
```{r}
p <- 3
g <- 3
m <- p*g * (g-1) * 0.5
total_phen <- pairwiseCI(data = dataset , total_phenols ~ cultivar, conf.level = 0.95/(2*m))
phen <- data.frame(Estimate = round(total_phen$byout[[1]]$estimate,2),
                   Lower = round(total_phen$byout[[1]]$lower,2),
                   Upper = round(total_phen$byout[[1]]$upper,2))
pander(phen,caption = "Confidence Intervals for Differences in Total Phenols")
#width <- pairwiseCI(data = data , width ~ sex, conf.level = 0.95/6)
#height <- pairwiseCI(data = data , height ~ sex, conf.level = 0.95/6)

phen$group <- c("2-1","3-1","3-2")


```
</center>

We see that the differences are such that it resembles the layout in the 3D scatterplot such that groups one and three are closer to group two (approximately one standard deviation away) than each other (approximately two standard deviations). The magnitudes are such that groups have the decreasing order as shown in the graph. Once again I can't say whether there is practical significance in the magnitude of these differences but given that the confidence intervals do not include zero these have statistical significance.

<center>
```{r,out.width= '80%'}

meanfl <- means %>% select(cultivar,flavanoids) %>% ggplot(aes(cultivar,flavanoids)) + 
  geom_point() + 
  geom_segment( aes(x=cultivar, xend=cultivar, y=0, yend=flavanoids)) + 
  theme_bw() + 
  labs(x = "Cultivar",y = "Mean Flavanoid Level",title = "Difference in Flavanoid Levels across Cultivars") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplotly(meanfl)
```
</center>
<center>
```{r}
flav <- pairwiseCI(data = dataset , flavanoids ~ cultivar, conf.level = 0.95/(2*m))
flav <- data.frame(Estimate = round(flav$byout[[1]]$estimate,2),
                   Lower = round(flav$byout[[1]]$lower,2),
                   Upper = round(flav$byout[[1]]$upper,2))
pander(flav,caption = "Confidence Intervals for Flavanoids")
```
</center>

A similar situation occurs for the amount of Flavanoids with the exception that the magnitudes are a bit difference though the overall orientation remains the same. It would seem that there is a larger difference between groups three and two compared to the difference between groups two and one. 

<center>
```{r,out.width= '80%'}

meanal <- means %>% select(cultivar,alcohol) %>% ggplot(aes(cultivar,alcohol)) + 
  geom_point() + 
  geom_segment( aes(x=cultivar, xend=cultivar, y=0, yend=alcohol)) + 
  theme_bw() + 
  labs(x = "Cultivar",y = "Mean Alcohol Level",title = "Difference in Alcohol Levels across Cultivars") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplotly(meanal)
```
</center>
<center>
```{r}
alco <- pairwiseCI(data = dataset , alcohol ~ cultivar, conf.level = 0.95/(2*m))
alco <- data.frame(Estimate = round(alco$byout[[1]]$estimate,2),
                   Lower = round(alco$byout[[1]]$lower,2),
                   Upper = round(alco$byout[[1]]$upper,2))
pander(alco,caption = "Confidence Intervals for Alcohol")

```
</center>

The difference structure still remains the same but it is now between different cultivars. The large difference now occurs between groups one and two instead of groups one and three.

## Conclusion

Overall we consider there to be real differences between the wine samples use different cultivars on all level pairs. Whether these differences are of practical significance is to be determined by someone who can interpret the magnitudes of the paired differences appropriately.  

For now we can say that there is a consistent structure to the differences in the cultivars across the total phenols and flavanoid variables where group one has the highest level followed by groups two and three respectively. 

However for the alcohol variable the order of magnitude is groups one, three then two.



## References

Johnson, R. A., & Wichern, D. W. (1992). Applied multivariate statistical analysis. Englewood Cliffs, N.J: Prentice Hall 

Lichman, M. (2013). UCI Machine Learning Repository <http://archive.ics.uci.edu/ml>. Irvine, CA: University of California, School of Information and Computer Science

NCSS Multivariate Analysis of Variance [pdf] Available at:
<https://ncss-wpengine.netdna-ssl.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Multivariate_Analysis_of_Variance-MANOVA.pdf>

Zhang J, Xu J (2009-jun). “On the k-sample Behrens-Fisher problem for high-dimensional data.” Science in China Series AMathematics, 52, 1285–1304.  